Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-07-25 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Draft

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Emperical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Furthermore, Natural forests are often comprised of
  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
  fires. Multiple studies have established the existence of non-additive
  effects in the flammability of dry litter, but we lack an understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption capacity and desorption rates. Burn trials were performed in moist
  litter beds of single species and of mixtures to establish flammability
  response to the dry-down process. Moisture dynamics vary across species with
  higher specific leaf area being associated with higher maximum absorption and
  lower litter bulk density being associated with faster drying rates. Species
  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
  \emph{Pinus} species) have faster rates of litter drying. We examined two
  axes of flammability variation (spread rate and fuel consumption) and found that these
  differing moisture dynamics result in time since wetting having strongly
  differing effects on flammability across the major litter types typified by
  different tree genera. Finally, we found that litter mixtures exhibited
  consistent non additivity in flame spread rate in fuel consumption. This non
  additivity became more positive with fuel drying. Litter mixtures burn with
  fire behavior more similar to that of the most flammable constituent species
  and this effect increases as fuels dry. }

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Environmental conditions such as temperature, relative
humidity, and wind drive the moisture content of the fuel available to burn
\cite{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
readily to environmental changes \cite{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \cite{Rothermel-1972, Nelson-2011}. Moisture acts as a
heat-sink resulting in less energy available for propagation, and the resulting
water vapor will both cool the flaming front and dilute flammable gases
\cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. Moisture of litter fuels varies with climate, with seasons and with
short term weather within a season. Therefore, the rate of drying and
flammability at any particular moisture content are key parameters that
influence fuel flammability throughout the fire season \cite{Nelson-2001}.

Fire scientists often categorize fuel by size classes that correspond to moisture exchange rates, although such simplified categories have long been recognized as not sufficient to describe species specific and decomposition effects on moisture dynamics \cite{Anderson-1985}. Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of models describing moisture dynamics of fine fuels are used by fire scientists \cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models usually aim to describe equilibrium conditions and at least  include temperature and relative humidity although some include precipitation and solar radiation effects as well. Explicit models of fire behavior as a function of time since wetting could improve fire prediction, however. In many litter fuel driven systems, natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event. If species differ in the rate of drying, this may strongly influence relatively flammability across species at any particular time post-wetting. Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires \cite{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}. However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be driven by by leaf traits, ermegent litter litter traits, or both. Species specific effects on litter flammability are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008} where species-specific litter drying rates play a role in flammability feedbacks \cite{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fueled by plant litter, overstory tree composition drives variation in local fire behavior \cite{Schwilk+Caprio-2011} as a result of differential litter flammability cross tree species and as a result of synergistic interactions in multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}. These flammability feedbacks may help explain historical fire frequency patterns, but current estimates of litter flammability in mixed conifer forests are based largely on burning dry litter.

Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye-2013}. Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated. After saturation, the differences in moisture content indicate different abilities to retain moisture. We propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density. Litter characteristics such as surface-area-to-volume ratio, litter bulk density, mass density, particle size and shape, and chemical traits influence moisture diffusion through litter [CITE]. Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits, especially leaf size \cite{Scarff+Westoby-2006}. Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate.

Flammability is a multi-dimensional characteristic of fuels \citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing different components of fire behavior. In litter driven fire, the two major axes of variation are described by measurements related to flame spread rate and measurements related to heat transfer. Flame spread rate is easy to measure in burning trials, but heat transfer can be more difficult and often proxies such as duration of heating are used \citep{Magalhaes+Schwilk-2012, Varner+Kane+etal-2015}. These different components of fire behavior have differing ecological effects. Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species. Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate. There is some evidence for non addiitvity of flammability in moist fuels and one study has found increasing non additivity with moisture \cite{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree species in a temperate fire-prone forest in California. We determined the maximum water retention and rate of moisture loss and examined how these vary in relation to two leaf and litter bed traits.  We then examined the influence of moisture on the flammability across these species and how flame spread rate and fuel consumption varied with time since wetting and by taxon. Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced two axes of flammability characterized by flame spread rate and total fuel consumed. Previous work described the flammability of these species in monocultures and in mixtures for oven-dry litter \cite{Magalhaes+Schwilk-2012} and found consistent positive non additivity of flammability where fire behavior of mixtures was driven by the most flammable species. Here, we examined if this positive non additivity was maintained in mixtures across varying levels of fuel moisture. The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects as seen in oven-dried litter. We tested if the non-additive effects on  reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest. We chose eight dominant tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws., \emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J. Buchholz. In this study, we consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed). Leaf litter was collected in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area. Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32\,km) for each species to control for plastic and ecotypic effects. An exception to this involved the litter from \emph{A. magnifica} which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints. Within that area, \emph{A. magnifica} litter was collected from over 10 different individuals. For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10\,m. The litter was collect about 2\,m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk. All litter was air dried down to < 5\% fuel moisture for weighing and dividing into moisture trials. The moisture manipulations and subsequent burning trials were conducted in April-August 2015.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets (approximately 45 x 45\,cm by 15\,cm
tall) constructed of 1 cm aluminum screen. Samples were 450 g of litter (dry
weight) per trial, and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples were
immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at 21\,C temperature and 30\% relative humidity, assessed via an iButton sensor
(Maxim Integrated) placed over the samples. Litter baskets were arranged in a
completely randomized design. During the dry-down process three subsamples were
taken at 24 hour intervals and weighed. These subsamples were then oven dried
for 24 hours at 100\,C and re-weighed to assess moisture content. This was
repeated until the final sample from a basket was less than 10\% moisture.

@@

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absoorption, specific
leaf area, and one litter trait known to directly effect flammability and
likely to influence drying rate, litter bulk density. Specific leaf area was
taken from leaf fresh leaf area and leaf dry mass from 5 individuals per
species and was collected as part of a previous study
\cite{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
experiment for each basket by dividing the dry mass (450\,g) by the estimated
litter volume (basket width x length x litter depth).

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these c oefficents we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$) for each species. All analyses were conducted in R
\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
Luke \cite{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that species within a genus tended to behave similarly and the two scale-leaved genera in the Cupressaceae,
\emph{Calocedrus} and \emph{Sequoiadendron}, also behaved similarly to one
another. Therefore, to preserve power and focus on analyses we used taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and treated these as fixed effects in our models. The effect of fuel moisture and
time since wetting on flammability parameters were modeled using linear
models with fuel moisture or time since wetting and taxon as fixed effects and
average vapor pressure deficit in the burning room during that trial as a
nuisance covariate. We used linear mixed models to predict spread rate, but
fuel consumption tends to exhibit threshold effects with few intermediate
values. Therefore, we used mixed generalized linear models with a beta
regression link with the `glmmADMB` package \cite{Skaug_Fournier_etal-2016} to
model percent fuel consumed as a function moisture content and of time since
wetting.

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped the
species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\cite{Lenth-2019}. This resulted in four broad groups of species and we then
selected one representative species from each litter type group and produced
all four possible three-species litter mixtures using those four species. For
each mixture and time since wetting, we predicted moisture content based on the
mean content of the three consituent species at that time since wetting. We
predicted flammability parameters based on the models fit to single species
data described above and then calculated an expected moisture content and
spread rate as the average of these three values for the mixture. The
expectation from the null model was that the difference between observed and
predicted values of flammability will be zero. We tested if there were
significant departures from the null (zero) with a mixed effects linear model.
For flammability parameters, we fit linear mixed effect models with the
observed-predicted flammability value as the dependent variable and used Wald
tests to test if the centered intercept was different than zero (overall non
additivity) and if the slope with time since wetting was different from zero
(non additivity changing with drying).

@@

There were marked differences in the extent of moisture species retain initially
following wetting and draining by gravity (all p < 0.0001, Fig.
\ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
broadleaf, is the species that retained the most moisture initially.
\emph{Quercus}, like the pines, had a rapid drying rate (Fig.
\ref{fig:drydown} and SI Fig. 1). Calculating estimated marginal means following the
linear mixed-effects model discriminated the species into three different
maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P.
  lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S.
  giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}
(SI Fig. 1). There are also differences in desiccation rates across the eight
species and estimated mixed linear model slopes discriminated species into
three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies
  spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
(SI Fig. 2).

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption{Dry down curves for eight litter types. Lines show best-fit
  exponential decay curves fit to moisture content on a dry mass basis. Litter
  samples we saturated by immersion for 24 hours and then allowed to drain by
  gravity for 3 min before initial weighing. Colors indicate genera and there
  were significant differences in both maximum water retention and in rate of drying (based linear model on log of water content). A small amount of horizontal jitter
  (0.5 hr) was added to this plot to aid in visualizing because, although
  collections were staggered, there were often multiple trays collected at a
  single time point.}
 \label{fig:drydown}
\end{figure}

\begin{table}[H]
  \caption{Linear mixed model results for moisture content as a function of time since
    wetting. Approximate degrees of freedom, pseudo F statistics and p-values
    were calculated by the Kenward-Roger approximation.
    \cite{Kenward_Roger-1997}.}
  \label{tab:drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\input{figs_tables/tab1_drydown_anova.ltx}
\end{table}

Across these eight species specific leaf area had a positive effect on maximum
water retention (Fig. \ref{fig:maxmc-di}, linear model p = 0.003) and litter
bulk density had a negative effect on desiccation rate (Fig. \ref{fig:bd-di},
linear model p = 0.029). Anova results are shown in Rables \ref{tab:mc_di_anova}.

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Emperical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Furthermore, Natural forests are often comprised of
  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
  fires. Multiple studies have established the existence of non-additive
  effects in the flammability of dry litter, but we lack an understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption capacity and desorption rates. Burn trials were performed in moist
  litter beds of single species and of mixtures to establish flammability
  response to the dry-down process. Moisture dynamics vary across species with
  higher specific leaf area being associated with higher maximum absorption and
  lower litter bulk density being associated with faster drying rates. Species
  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
  \emph{Pinus} species) have faster rates of litter drying. We examined two
  axes of flammability variation (spread rate and fuel consumption) and found that these
  differing moisture dynamics result in time since wetting having strongly
  differing effects on flammability across the major litter types typified by
  different tree genera. Finally, we found that litter mixtures exhibited
  consistent non additivity in flame spread rate in fuel consumption. This non
  additivity became more positive with fuel drying. Litter mixtures burn with
  fire behavior more similar to that of the most flammable constituent species
  and this effect increases as fuels dry. ADD Synthesis statement.}

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Environmental conditions such as temperature, relative
humidity, and wind drive the moisture content of the fuel available to burn
\cite{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
readily to environmental changes \cite{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \cite{Rothermel-1972, Nelson-2011}. Moisture acts as a
heat-sink resulting in less energy available for propagation, and the resulting
water vapor will both cool the flaming front and dilute flammable gases
\cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. Moisture of litter fuels varies with climate, with seasons, and with
short term weather within a season. Therefore, the rate of drying and
flammability at any particular moisture content are key parameters that
influence fuel flammability throughout the fire season \cite{Nelson-2001}.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \cite{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuelbeds
\cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. In many litter fuel driven
systems, natural lightning ignitions are concurrent with precipitation and,
therefore, fire behavior can depend on rate of drying following a wetting
event. If species differ in their drying rate, this may strongly influence
relatively flammability across species at any particular time following wetting.
Such differential moisture content across time, may contribute to the often
high spatial heterogeneity of surface fires \cite{Knapp_Schwilk_etal-2006,
  Kreye_Hiers_etal-2018}.

% Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936,   Fons-1946, Anderson+Rothermal-1965}.

Litter packing (measured as litterbed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability, especially on the flame spread rate axis.
Leaf litter density is driven by leaf size and shape
\cite{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \cite{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established, however, how moisture dynamics modulate species specific
flammability is poorly known, and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \cite{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \cite{Schwilk+Caprio-2011} as a result of differential litter
flammability cross tree species and as a result of synergistic interactions in
multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}. These flammability
feedbacks may help explain historical fire frequency patterns, but current
estimates of litter flammability in mixed conifer forests are based largely on
burning dry litter.

Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}. Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated. After saturation, the differences in moisture content indicate different abilities to retain moisture. Litter particle geometry may have direct effects on maximum moisture absorption through area-mass relationships and we also expect effects of leaf morphaology, espcially leaf cuticle thickness \cite{Van_Wagner-1969}. Drying rates should also be influenced by leaf and litter particle traits: surface-area-to-volume ratio and litter bulk density will influence air flow through the litter bed. Litter packing therefore has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits, especially leaf size \cite{Scarff+Westoby-2006}. Fuelbeds composed of larger leaves with lower moisture retainability and should favor drying, thus increasing the probability of ignition and fire spread.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer. Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects. Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects have been shown to be synergistic in dry
litter fuels where the flammability of a mixture is driven by the most
flammable species \cite{VanAltena+Logtestjin+etal-2012,
  Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
influence litter moisture content, especially if different leaf traits
influence adsorptive capacity and drying rate. There is some evidence for
non-additive flammability in moist fuels with one study finding increasing
non-additivity with moisture \cite{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits. We then examined the influence
of moisture on the flammability across these species. Experimental
burn trials of leaf litter at various moisture levels determined how time since
wetting influenced two axes of flammability characterized by flame spread rate
and total fuel consumed. Previous work described the flammability of these
species in monocultures and in mixtures for oven-dry litter
\cite{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
flammability where fire behavior of mixtures was driven by the most flammable
species. Here, we examined if this positive non additivity was maintained in
mixtures across varying levels of fuel moisture. The burn trials were also
performed with multi-species litter mixtures to evaluate if species contribute
equally to litter moisture dynamics or if there are non-additive effects as
observed in oven-dried litter. We tested if the non-additive effects on reported in
previous studies occur under ecologically-relevant moisture levels and whether
fuel moisture effects strengthen or mask such non-additivity.

@@

Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest. We chose eight dominant tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws., \emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J. Buchholz. In this study, we consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed). Leaf litter was collected in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area. We collected from a minimum of four populations per species with each population 9--32\,km distant from its neighbors. An exception to this was \emph{A. magnifica} which is restricted to the higher elevations and could only be collected from one area, due to designated wilderness boundaries and logistical constraints. Within that area, \emph{A. magnifica} litter was collected from over 10 different individuals. For the remaining species, the collection involved 2--4 individual trees at each site, separated from one another by at least 10\,m. The litter was collected 2\,m away from the trunk to obtain a more uniform sample of leaf and fine branchlet litter, because bark and heavier branches tend to fall closer to the tree trunk. All litter was air dried to < 5\% fuel moisture by dry weight for weighing and dividing into moisture trials. The moisture manipulations and subsequent burning trials were conducted in April--August 2015.

@@

We produced dry-down curves to determine maximum water retainability and the
rate of moisture content loss for each species and for four distinct
three-species mixtures. These curves allowed us to determine the rate at which
a species loses moisture, approximated as exponential decay, and to determine
maximum water content following saturation and gravity draining. Samples from
the eight species (six replicates per species) and four mixtures (5 replicates
per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
tall constructed of 1\,cm aluminum screen. Samples were 450\,g dry weight
litter per trial and litter depth was standardized to 10\,cm by changing the
horizontal dimensions of the basket. Weight was measured by using a balance
sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
were immersed in plastic storage bins filled with water for 24 hours to fully
saturate them. After saturation, the baskets were removed from the water,
allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
content. The samples were then allowed to dry in a controlled environment kept
at 21\,C temperature and 30\% relative humidity, assessed via an iButton sensor
(Maxim Integrated) placed over the samples. Litter baskets were arranged in a
completely randomized design. During the dry-down process three subsamples were
taken at 24 hour intervals and weighed. These subsamples were then oven dried
for 24 hours at 100\,C and re-weighed to assess moisture content. This was
repeated until the final sample from a basket was less than 10\% moisture.

@@

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption, specific
leaf area, and one litter trait known to directly effect flammability and
likely to influence drying rate, litter bulk density. Specific leaf area was
taken from leaf fresh leaf area and leaf dry mass from 5 individuals per
species and was collected as part of a previous study
\cite{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
experiment for each basket by dividing the dry mass (450\,g) by the estimated
litter volume (basket width x length x litter depth).

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficents we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
Luke \cite{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that species within a genus tended to behave similarly and the two scale-leaved genera in the Cupressaceae,
\emph{Calocedrus} and \emph{Sequoiadendron}, also behaved similarly to one
another. Therefore, to preserve power used taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and treated these as fixed effects in our models. The effect of fuel moisture and
time since wetting on flammability parameters were modeled using linear
models with fuel moisture or time since wetting and taxon as fixed effects and
average vapor pressure deficit in the burning room during that trial as a
nuisance covariate. We used linear mixed models to predict spread rate, but
fuel consumption tends to exhibit threshold effects with few intermediate
values. Therefore, we used mixed generalized linear models with a beta
regression link with the `glmmADMB` package \cite{Skaug_Fournier_etal-2016} to
model percent fuel consumed as a function moisture content and of time since
wetting.

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped
the species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\cite{Lenth-2019}. This resulted in four groups of species and we then selected
one representative species from each litter type group and produced all four
possible three-species litter mixtures using those four species. For each
mixture and time since wetting, we predicted moisture content based on the mean
content of the three consituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null (zero) with a mixed effects
linear model. For flammability parameters, we fit linear mixed effect models
with the observed-predicted flammability value as the dependent variable and
used Wald tests to test if the centered intercept was different than zero (a
measure of overall non-additivity) and if the slope with time since wetting was
different from zero (indicating non-additivity changing with drying).

@@

There were marked differences in the extent of moisture species retain initially
following wetting and draining by gravity (all $P$ < 0.0001, Fig.
\ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
broadleaf, is the species that retained the most moisture initially.
\emph{Quercus}, like the pines, had a rapid drying rate (Fig.
\ref{fig:drydown} and SI Fig. 1). Calculating estimated marginal means following the
linear mixed-effects model discriminated the species into three different
maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P.
  lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S.
  giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}
(SI Fig. 1). There are also differences in desiccation rates across the eight
species and estimated mixed linear model slopes discriminated species into
three groups from lowest to most rapid: 1) \emph{C.decurrens}, 2) \emph{Abies
  spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
(SI Fig. 2).

\begin{figure}[H]
  \centering
\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
\caption{Dry down curves for eight litter types. Lines show best-fit
  exponential decay curves fit to moisture content on a dry mass basis. Litter
  samples we saturated by immersion for 24 hours and then allowed to drain by
  gravity for 3 min before initial weighing. Colors indicate genera and there
  were significant differences in both maximum water retention and in rate of drying (based linear model on log of water content). A small amount of horizontal jitter
  (0.5 hr) was added to this plot to aid in visualizing because, although
  collections were staggered, there were sometimes multiple trays collected at a
  single time point.}
 \label{fig:drydown}
\end{figure}

\begin{table}[H]
  \caption{Linear mixed model results for moisture content as a function of time since
    wetting. Approximate degrees of freedom (df), pseudo F statistics and p-values
    were calculated by the Kenward-Roger approximation.
    \cite{Kenward_Roger-1997}.}
  \label{tab:drydown}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{llll}
  \toprule
Effect & df & $F$ & $P$ \\ 
  \midrule
hour & 1, 183.73 & 1599.59 & $<$.0001 \\ 
  spcode & 7, 89.73 & 36.15 & $<$.0001 \\ 
  hour:spcode & 7, 184.15 & 23.06 & $<$.0001 \\ 
   \bottomrule
\end{tabular}
\end{table}
